---
output: github_document
---

<!-- README.md is generated from README.Rmd. Please edit that file -->

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.path = "man/figures/README-",
  out.width = "100%"
)
```

# legaciesr

<!-- <!-- badges: start --> -->

<!-- [![Codecov test coverage](https://codecov.io/gh/ccappelen/legaciesR/graph/badge.svg)](https://app.codecov.io/gh/ccappelen/legaciesR) -->
<!-- [![R-CMD-check](https://github.com/ccappelen/legaciesR/actions/workflows/R-CMD-check.yaml/badge.svg)](https://github.com/ccappelen/legaciesR/actions/workflows/R-CMD-check.yaml) -->
<!-- <!-- badges: end --> -->

The `legaciesr` package provides a set of functions used to aggregate and summarize the mapping data collected in the [LEGACIES](https://www.legacies-project.com) project. It allows users to, i.a., (1) create contour polygons capturing the territorial extent of historical states at different probability thresholds, (2) create a grid with various summary measures of historical state presence, and (3) add a range of commonly used covariates to the polygon and grid data.

NB: The package is still experimental and no functions should be considered stable.

## Installation

You can install the development version of legaciesr from [GitHub](https://github.com/) with:

``` r
# install.packages("devtools")
devtools::install_github("ccappelen/legaciesR")
```


```{r packages, eval = TRUE, message = FALSE}
library(legaciesr) 
library(sf) 
library(dplyr)
```

## Errors in current master_shapefile
Currently, one geometry (COWID = "EGY") fails to rebuild due to crossing edges. However, a few other COWIDs result in invalid geometries when creating contour polygons. For now, these are omitted:

```{r omit COWIDs with invalid geometries, eval = FALSE}
df <- df %>% 
  filter(!COWID %in% c("EGY", "LUN", "KUN", "SAF"))
```


## Handling of invalid geometries

The functions in `legaciesr` will (by default) attempt to fix invalid geometries which would otherwise return an error. The fixes rely on `legaciesr::fix_invalid()` which is a wrapper around the `sf::st_make_valid()` function that attempts to reiteratively lower the allowed precision to rebuild valid geometries. However, because invalid geometries can happen for many reasons that may point to other data errors as well, it is recommended to check for valid geometries and run the `legaciesr::fix_invalid()` to identify potential issues. See package documentation for further information on the use of this function.

```{r fix invalid geometries, eval = TRUE}
df <- example_df # Load example data set (based on random data)
df <- fix_invalid(df)

```


## Create contour polygons for a single group

See documentation for additional arguments (e.g., threshold for number of maps, the number of contour polygons, resolution of raster, and handling of invalid geometries).

```{r contour polygons for single group, eval = TRUE}
df1 <- df[df$name == unique(df$name)[8], ]
contour_polygons(df1, id_vars = name)
```

## Create contour polygons across all groups

See documentation for additional arguments (e.g., specifying the period intervals to group by, thresholds for number of maps within a group, etc.).

```{r contour polygons across all groups, eval = TRUE}
get_contours(df, grp_id = name)
```

## Create contour polygons across all groups and by period

```{r contour polygons across COWIDs and period, eval = TRUE}
get_contours(df, by_period = T, grp_id = name, period_id = year)
```

## Support for parallel processing:

Parallel processing is is implemented using the [future::future] framework. There are two ways of running jobs in parallel: `multicore` which uses 'forking' to run multiple jobs in parallel with shared memory and `multisession` which launches a set of background R sessions. 'Forking' can be faster than multisession because of the larger overhead associated with copying the active environment to each background R session (whereas forking processes shares memory). However, 'forking' is not supported on Windows platforms and is considered unstable when running from within RStudio (on both Windows and Unix systems such as MacOS). The function will automatically determine whether `multicore` is supported by the platform and choose the appropriate plan.

The greater overhead associated with `multisession` is primarily during the first parallel run in a given R session (since the background R sessions stays available for additional parallel jobs). It is possible to define a [future::plan("multisession")] in the global environment, which will minimize overhead in subsequent parallel jobs (apart from the first). The function will automatically detect if a `multisession` plan has been set globally and, thus, will not close background sessions after running.
