---
output: github_document
---

<!-- README.md is generated from README.Rmd. Please edit that file -->

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.path = "man/figures/README-",
  out.width = "100%"
)
```

# legaciesr

<!-- badges: start -->

<!-- badges: end -->

The `legaciesr` package provides a set of functions used to aggregate and summarize the mapping data collected in the [LEGACIES](https://www.legacies-project.com) project. It allows users to, i.a., (1) create contour polygons capturing the territorial extent of historical states at different probability thresholds, (2) create a grid with various summary measures of historical state presence, and (3) add a range of commonly used covariates to the polygon and grid data.

NB: The package is still experimental and no functions should be considered stable.

## Installation

You can install the development version of legaciesr from [GitHub](https://github.com/) with:

``` r
# install.packages("devtools")
devtools::install_github("ccappelen/legaciesR")
```

```{r packages, eval = TRUE, message = FALSE}
library(legaciesr) 
library(sf) 
library(dplyr)
```

## Handling of invalid geometries

The functions in `legaciesr` will (by default) attempt to fix invalid geometries which would otherwise return an error. The fixes rely on `legaciesr::fix_invalid()` which is a wrapper around the `sf::st_make_valid()` function that attempts to reiteratively lower the allowed precision to rebuild valid geometries. However, because invalid geometries can happen for many reasons that may point to other data errors as well, it is recommended to check for valid geometries and run the `legaciesr::fix_invalid()` to identify potential issues. See package documentation for further information on the use of this function.

```{r fix invalid geometries, eval = TRUE}
df_path <- file.path("../../LEGACIES/Test code/test_shape") # PATH FOR FOLDER OF master_shapefile (user-defined)
df <- read_sf(df_path, "master_shapefile") # LOAD master_shapefile

df <- fix_invalid(df)

```

Currently, one geometry (COWID = "EGY") fails to rebuild due to crossing edges. However, a few other COWIDs result in invalid geometries when creating contour polygons. For now, these are omitted:

```{r omit COWIDs with invalid geometries, eval = FALSE}
df <- df %>% 
  filter(!COWID %in% c("EGY", "LUN", "KUN", "SAF"))
```

## Create contour polygons for a single COWID

See documentation for additional arguments (e.g., threshold for number of maps, the number of contour polygons, resolution of raster, and handling of invalid geometries).

```{r contour polygons for single COWID, eval = FALSE}
sok <- df %>%
  filter(COWID == "SOK")
contour_polygons(sok, id_vars = COWID)
```

## Create contour polygons across all COWIDs

See documentation for additional arguments (e.g., specifying the period intervals to group by, thresholds for number of maps within a group, etc.).

```{r contour polygons across all COWIDs, eval = FALSE}
get_contours(df, grp_id = COWID)
```

## Create contour polygons across all COWIDs and by period

```{r contour polygons across COWIDs and period, eval = FALSE}
get_contours(df, by_period = T, grp_id = COWID, period_id = year)
```

## Support for parallel processing:

There are two ways of running jobs in parallel. Forked R processes or running multiple background R sessions. In the current setup, running multiple background processes (multisession) are slower than running the jobs sequentially due to the overhead associated with opening new R sessions. However, machines running on Microsoft Windows do not support forking (multicore) and will therefore default to a sequential plan unless `parallel` is set to TRUE. On Unix platforms (e.g., MacOS), it will default to parallel processing.
